Formal Modeling and Verification of a Rate-Monotonic Scheduling Implementation with Real-Time Maude

### 2015-11-05 RMS/submission.tex
changed to non-compsoc style, and modified until introducing real-time maude

\documentclass[10pt,journal,compsoc]{IEEEtran}
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[10pt,journal,compsoc]{../sty/IEEEtran}

@@

\author{Jiaxiang~Liu,
        Min~Zhou,
        Xiaoyu~Song,
        Ming~Gu,
        and~Jiaguang~Sun% <-this % stops a space
\IEEEcompsocitemizethanks{%
\IEEEcompsocthanksitem J. Liu is with the School of Software, Tsinghua
University, Beijing 100084, China, and also with LIX, \'Ecole
Polytechnique, Palaiseau 91120, France.\protect\\ E-mail:
jiaxiang.liu@hotmail.com
\IEEEcompsocthanksitem M. Zhou, M. Gu and J. Sun are with the School
of Software, Tsinghua University, Beijing 100084, China.
\IEEEcompsocthanksitem X. Song is with the Department of Electrical \&
Computer Engineering, Portland State University, Portland, OR
97207-0751, USA.}% <-this % stops an unwanted space
}
%\thanks{Manuscript received April 19, 2005; revised September 17, 2014.}}

% note the % following the last \IEEEmembership and also \thanks - 
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
% 
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
% "AB" then you have to do: "\textbf{A}\textbf{B}"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.

@@

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Computer Society, IEEEtran, journal, \LaTeX, paper, template. TODO.
\end{IEEEkeywords}}

@@

\IEEEraisesectionheading{\section{Introduction}\label{s:introduction}}
% Computer Society journal (but not conference!) papers do something unusual
% with the very first section heading (almost always called "Introduction").
% They place it ABOVE the main text! IEEEtran.cls does not automatically do
% this for you, but you can achieve this effect with the provided
% \IEEEraisesectionheading{} command. Note the need to keep any \label that
% is to refer to the section immediately after \section in the above as
% \IEEEraisesectionheading puts \section within a raised box.

% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps (small caps for compsoc).
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{P}{eriodic} task scheduling is one of the most important
topics within the field of real-time systems, due to the large number
of control systems that require cyclic
activities~\cite{buttazzo2011hard}. A set of periodic tasks is said to
be \emph{schedulable} with respect to some scheduling algorithm if all
jobs meet their deadlines. \emph{Rate-Monotonic Scheduling}
(\emph{RMS}) is a \emph{fixed} priority scheduling algorithm for
preemptive hard real-time environments proposed by Liu and
Layland~\cite{DBLP:journals/jacm/LiuL73}, which assigns priorities to
jobs according to the periods of the corresponding tasks: the smaller
period, the higher priority. RMS is proved to be the \emph{optimal}
fixed priority scheduling algorithm~\cite{DBLP:journals/jacm/LiuL73},
in the sense that any set of tasks, which is schedulable under
\emph{some} fixed priority scheduling algorithm, is also schedulable
with respect to RMS. It is widely used in safety-critical real-time
applications, such as trains and avionics, thanks to its optimality
and easiness to implement.

@@

Extensive effort to apply formal methods, such as model checking and
theorem proving, has been made to analyze safety-critical systems for
the past few
years~\cite{DBLP:journals/iandc/MeseguerR13,DBLP:journals/cacm/Leroy09,DBLP:conf/sosp/KleinEHACDEEKNSTW09}. However,
as far as we know, few~\cite{DBLP:conf/iceccs/CuiDT14,TianD2011}
attempt to analyze the RMS algorithm, while no work for
implementations of RMS is found.

In this paper, we use \emph{Real-Time Maude}, a \emph{rewriting}-based
modeling language and analyzing tool for real-time systems, to model a
realistic implementation of RMS and then verify some desired
properties on the built model. Based on a realistic implementation,
our model extends the standard one proposed
in~\cite{DBLP:journals/jacm/LiuL73}, by allowing computation
requirements of tasks to be non-constant, and by considering overhead
and other details of the environments.

The rest of this paper is organized as
follows. Section~\ref{s:background} gives some background of the RMS
algorithm and Real-Time Maude.  Section~\ref{s:imp} presents the RMS
implementation that we model and analyze.  Section~\ref{s:formalism}
introduces how we model the RMS implementation using Real-Time
Maude. Then Section~\ref{s:verification} explains how to verify the
desired properties and to evaluate the results. Related work is
discussed in Section~\ref{s:relate}. We conclude the paper in
Section~\ref{s:conclusion}.

@@

\subsection{Rate-Monotonic Scheduling Algorithm}
\label{ss:rms}
%\subsection{Formulation of the Standard Setting}
A task set consists of \emph{only} $n$ periodic tasks
$\tau_1,\ldots,\tau_n$. Each task $\tau_i$ has a period $T_i$ and a
computation requirement $C_i$. First jobs of all tasks are assumed to
be initiated simultaneously. This means that jobs corresponding to
task $\tau_i$ are initiated at times $kT_i$ with integers $k\ge
0$. Deadlines consist of runnability constraints only: the job
initiated at time $kT_i$ has $(k+1)T_i$ as its deadline, which is the
initiation time of the next job. The RMS algorithm chooses the
labeling such that $T_1\le T_2\le \ldots \le T_n$. Consequently,
$\tau_i$ receives priority $i$, assuming smaller numbers have higher
priorities. The following assumptions are made:

(A1) Computation requirement $C_i$ for each task $\tau_i$ is constant
and does not vary with time.

(A2) Tasks are independent, such that they are ready to run at their
initiation times and can be preempted instantly (ignoring all
blocking).

(A3) All overhead, such as task switching times, is ignored.

However in this paper, we consider an implementation instead of the
RMS algorithm itself, thus the model would be more complicated than
this standard, ideal setting.  Assumptions~(A1) and (A3) will be
relaxed later to obtain a more realistic analysis model.

\subsection{Real-Time Maude}
Based on \emph{rewriting logic}~\cite{DBLP:journals/jlp/Meseguer12},
\emph{Real-Time Maude} is a language and tool that extends
\emph{Maude} to support the formal specification and analysis of
real-time systems~\cite{DBLP:journals/lisp/OlveczkyM07}.

\subsubsection{Specification}
Real-Time Maude models systems as \emph{modules}. A module specifies a
\emph{real-time rewrite theory} ${\cal R} = (\Sigma, E\cup A , IR,
TR)$, where:
\begin{itemize}
\item $\Sigma$ is an algebraic \emph{signature}, that is, a set of
  declarations of \emph{sorts}, \emph{subsorts} and \emph{function
    symbols}. The function symbols are allowed to be mixfix.
\item $(\Sigma, E\cup A)$ is a \emph{membership equational logic
  theory}, with $E$ a set of conditional equations on $\Sigma$, and
  $A$ a set of equational axioms such as associativity, commutativity
  and identity.  $(\Sigma, E\cup A)$ models the system's ``static''
  states as an algebraic data type, and includes a built-in
  specification of a sort \verb|Time|.
\item $IR$ is a set of \emph{labeled conditional rewrite rules}
  specifying the system's local transitions. Each rule has the form
  $[l]~:~t\rightarrow t'\mbox{ \textbf{if} }\bigwedge^n_{j=1}cond_j$,
  where each $cond_j$ is an equality $u_j=v_j$ and $l$ is a
  \emph{label}. Such a rule specifies an \emph{instantaneous
    transition}, without consuming time, from an instance of $t$ to
  the corresponding instance of $t'$, \emph{provided} the conditions
  hold.
\item $TR$ is a set of (\emph{labeled}) \emph{tick rules}
  $[l]~:~\{t\}\rightarrow\{t'\} \mbox{ \textbf{in time} }r\mbox{
  \textbf{if} }cond$ that specify \emph{timed transitions} advancing
  time in the \emph{entire} state $t$ by $r$ time units. $IR$ and $TR$
  together model the ``dynamic'' behaviors of the system.
\end{itemize}

>>>>

\documentclass[10pt,journal]{IEEEtran}
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[10pt,journal,compsoc]{../sty/IEEEtran}

@@

\hide{
\author{Jiaxiang~Liu,
        Min~Zhou,
        Xiaoyu~Song,
        Ming~Gu,
        and~Jiaguang~Sun% <-this % stops a space
\IEEEcompsocitemizethanks{%
\IEEEcompsocthanksitem J. Liu is with the School of Software, Tsinghua
University, Beijing 100084, China, and also with LIX, \'Ecole
Polytechnique, Palaiseau 91120, France.\protect\\ E-mail:
jiaxiang.liu@hotmail.com
\IEEEcompsocthanksitem M. Zhou, M. Gu and J. Sun are with the School
of Software, Tsinghua University, Beijing 100084, China.
\IEEEcompsocthanksitem X. Song is with the Department of Electrical \&
Computer Engineering, Portland State University, Portland, OR
97207-0751, USA.}% <-this % stops an unwanted space
}
%\thanks{Manuscript received April 19, 2005; revised September 17, 2014.}}

% note the % following the last \IEEEmembership and also \thanks - 
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
% 
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
% "AB" then you have to do: "\textbf{A}\textbf{B}"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.
}

@@

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
  Real-time systems, scheduling, embedded systems, modeling, formal
  verification, rewriting logic.
\end{IEEEkeywords}
}

@@

%\IEEEraisesectionheading{\section{Introduction}\label{s:introduction}}
\section{Introduction}\label{s:introduction}
% Computer Society journal (but not conference!) papers do something unusual
% with the very first section heading (almost always called "Introduction").
% They place it ABOVE the main text! IEEEtran.cls does not automatically do
% this for you, but you can achieve this effect with the provided
% \IEEEraisesectionheading{} command. Note the need to keep any \label that
% is to refer to the section immediately after \section in the above as
% \IEEEraisesectionheading puts \section within a raised box.

% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps (small caps for compsoc).
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{P}{eriodic} task scheduling is one of the most important
topics within the field of real-time systems, due to the large number
of control systems that require cyclic
activities~\cite{buttazzo2011hard}. A set of periodic tasks is said to
be \emph{schedulable} with respect to some scheduling algorithm if all
jobs meet their deadlines. \emph{Rate-Monotonic Scheduling}
(\emph{RMS}) is a \emph{fixed} priority scheduling algorithm for
preemptive hard real-time environments proposed by Liu and
Layland~\cite{DBLP:journals/jacm/LiuL73}, which assigns priorities to
jobs according to the periods of the corresponding tasks: the smaller
period, the higher priority. RMS is proved to be the \emph{optimal}
fixed priority scheduling algorithm~\cite{DBLP:journals/jacm/LiuL73},
in the sense that any set of tasks, which is schedulable under
\emph{some} fixed priority scheduling algorithm, is also schedulable
with respect to RMS. It is widely used in safety-critical real-time
applications, such as vehicles and avionics, thanks to its optimality
and easiness to implement.

@@

It is also crucial to ensure the reliability of the implementation of
RMS when it serves in a safety-critical system. When analyzing a
realistic implementation instead of the algorithm, theoretical results
may be no more applicable. For instance, switching overhead may lead
to unschedulability even if the task set satisfies the schedulability
condition derived from algorithm analysis. Extensive effort to apply
formal methods, such as model checking and theorem proving, has been
made to analyze safety-critical systems for the past few
years~\cite{DBLP:journals/tie/MiyawakiMSYV05,DBLP:journals/iandc/MeseguerR13,DBLP:journals/cacm/Leroy09,DBLP:conf/sosp/KleinEHACDEEKNSTW09}. However,
as far as we know, few~\cite{DBLP:conf/iceccs/CuiDT14,TianD2011}
attempt to analyze the RMS algorithm, while no work for
implementations of RMS is found.

In this paper, we use \emph{Real-Time Maude}, a \emph{rewriting}-based
formal modeling language and analyzing tool for real-time systems, to
model a realistic implementation of RMS, which serves as a simplified
operating system within an avionic control system, and then verify
several desired properties. Based on a realistic implementation, our
model extends the standard RMS model proposed
in~\cite{DBLP:journals/jacm/LiuL73}, by allowing computation
requirements of tasks to be non-constant, and by considering overhead
and other details of the hardware platform.

The rest of this paper is organized as
follows. Section~\ref{s:background} gives some background of both the
RMS algorithm and Real-Time Maude.  Section~\ref{s:imp} presents the
RMS implementation that we model and analyze.
Section~\ref{s:formalism} introduces how we model the RMS
implementation using Real-Time Maude. Then
Section~\ref{s:verification} explains how to verify the desired
properties and to evaluate the results. Related work is discussed in
Section~\ref{s:relate}. We conclude the paper in
Section~\ref{s:conclusion}.

@@

\subsection{Rate-Monotonic Scheduling Algorithm}
\label{ss:rms}
%\subsection{Formulation of the Standard Setting}
A task set consists of \emph{only} $n$ periodic tasks
$\tau_1,\ldots,\tau_n$. Each task $\tau_i$ has a period $T_i$ and a
computation requirement $C_i$. First jobs of all tasks are assumed to
be initiated at time $0$ simultaneously.  Deadlines consist of
runnability constraints only: the deadline of a job corresponding to
$\tau_i$ is the initiation time of the next job corresponding to
$\tau_i$.  The RMS algorithm chooses the labeling such that $T_1\le
T_2\le \ldots \le T_n$. Consequently, $\tau_i$ receives priority $i$,
assuming smaller numbers have higher priorities. The following
assumptions are made:

(A1) Jobs corresponding to task $\tau_i$ are initiated exactly at
times $kT_i$ with integers $k\ge 0$.

(A2) Computation requirement $C_i$ for each task $\tau_i$ is constant
and does not vary with time.

(A3) Tasks are independent, such that they are ready to run at their
initiation times and can be preempted instantly (ignoring all
blocking).

(A4) All overhead, such as task switching times, is ignored.

However in this paper, we consider an implementation instead of the
RMS algorithm itself, thus the model would be more complicated than
this standard, ideal setting. Assumption~(A1) will be modified because
of the interrupt masking mechanism, while (A2) and (A4) are relaxed to
obtain a more realistic analysis model.

\subsection{Real-Time Maude}
\emph{Real-Time Maude}~\cite{DBLP:journals/lisp/OlveczkyM07} is an
extension of \emph{Maude}~\cite{DBLP:journals/tcs/ClavelDELMMQ02}
which is a language and tool based on \emph{rewriting
  logic}~\cite{DBLP:journals/jlp/Meseguer12}.  It supports formal
specification and analysis of real-time systems.

\subsubsection{Specification}
Real-Time Maude models systems as \emph{modules}. A module specifies a
\emph{real-time rewrite theory} ${\cal R} = (\Sigma, E\cup A , IR,
TR)$, where:
\begin{itemize}
\item $\Sigma$ is an algebraic \emph{signature}, that is, a set of
  declarations of \emph{sorts}, \emph{subsorts} and \emph{function
    symbols}. The function symbols are allowed to be mixfix.
\item $(\Sigma, E\cup A)$ is a \emph{membership equational logic
  theory}, with $E$ a set of conditional \emph{equations} and
  \emph{memberships} on $\Sigma$, and $A$ a set of equational axioms
  such as associativity, commutativity and identity.  $(\Sigma, E\cup
  A)$ models the system's ``static'' states as an algebraic data type,
  and is equipped with a built-in specification of a sort \verb|Time|.
\item $IR$ is a set of \emph{labeled conditional rewrite rules}
  specifying the system's local transitions. Each rule has the form
  $[l]~:~s\rightarrow s'\mbox{ \textbf{if} }\bigwedge^n_{j=1}cond_j$,
  where each $cond_j$ is an equality $u_j=v_j$ and $l$ is a
  \emph{label}. Such a rule specifies an \emph{instantaneous
    transition}, without consuming time, from an instance of $s$ to
  the corresponding instance of $s'$, \emph{provided} the conditions
  hold.
\item $TR$ is a set of (\emph{labeled}) \emph{tick rules} of the form
  $[l]~:~\{s\}\rightarrow\{s'\} \mbox{ \textbf{in time} }t\mbox{
  \textbf{if} }cond$ that specify \emph{timed transitions}, which
  advances time in the \emph{entire} state $s$ by $t$ time units. $IR$
  and $TR$ together model the ``dynamic'' behaviors of the system.
\end{itemize}
